---
title: "Morphological phylogeny with `phangorn`"
author: 
- name: Iris Bardel-Kahr
  affiliation: University of Graz
  email: iris.bardel-kahr@uni-graz.at
output: html_document
bibliography: references_mm.bib
date: "Last compiled on `r format(Sys.time(), '%d %B, %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("Z:/Pfingstl-Kerschbaumer/Genetik_Iris/Morpho_Matrix/")
#knitr::opts_chunk$set(fig.width=12, fig.height=8) #hope this works...
```

## Load packages

The package `phangorn` [@phangorn] (with the dependency `ape` [@ape]) was used for this analyses (starting from a **nexus** file). If you start with a **.csv** file, `dplyr` [@dplyr] and `tidyr` [@tidyr] can be used for data manipulation.  
`set.seed` is used to get reproducible results.
  
```{r packages, message=F}
#install.packages("ape")
#install.packages("phangorn")

# for the latest development version, you can use:
#remotes::install_github("KlausVigo/phangorn")

#install.packages("tidyr")
#install.packages("dplyr")
library(ape)
library(phangorn)
library(tidyr)
library(dplyr)
set.seed(102)
```
\

## Load data

#### from a **.csv** file

```{r load_csv}
morpho_data <- read.csv("Matrix.csv")
```
\

After reading in the **.csv** file, the data has to be changed into the right format to be read as `phyDat` by `phangorn`.
  
  
```{r change_csv}
morpho_data$species <- gsub(" ", "_", morpho_data$species)

morpho_data <- morpho_data %>% 
  mutate_if(is.numeric, as.character) %>%
  pivot_longer(cols = -species, names_to = "trait", values_to = "value") %>%
  pivot_wider(names_from = species, values_from = value) %>%
  select(-trait)
```
\
  
#### from a **.nex** (nexus) file

```{r read_nex}
morpho_data <- read.nexus.data("morpho_matrix_full_names.nex")
```
\
  
#### `phyDat` format for subsequent analyses

For morphometric data, `type="USER"` needs to be selected and a vector reflecting the data has to be given to `levels`. Here we have the states 0 to 8.

```{r phydat}
morpho_pd <- phyDat(morpho_data, type = "USER", levels=as.character(0:8), ambiguity="?")
```
\

## Parsimony Trees

`pratchet` from *phangorn* uses the parsimony ratchet [@nixon1999parsimony]. `all=TRUE` gives back all possible trees with the same parsimony score. `k` (minimum number of iterations when no change in the parsimony score is detected) around **50** is sufficient for this analysis. For unordered data `method="fitch"` (the default) is much quicker and gives back the same results as `method="sankoff"`.
The following `acctran` assigns branch lengths to the trees.

```{r parsimony, message=F, cache=TRUE}
tree_pars <- pratchet(morpho_pd, k = 50, minit = 1000,
                     maxit = 100000, all=TRUE, trace = 0)
tree_pars <- acctran(tree_pars, morpho_pd)
tree_pars
```
\

There are **14** trees with lowest parsimony score.  
The outgroup is *Collohmannia johnstoni*, so we root the trees accordingly.

```{r root}
rooted_tree <- root(tree_pars, "Collohmannia_johnstoni", resolve.root=TRUE, edgelabel=TRUE)
```
\

To balance out the root, we write a function and apply it to all trees in the `multiPhylo` object.

```{r pretty_root}
pretty_root <- function(x){
  root <-  getRoot(x)
  ind <- which(x$edge[,1] == root)
  el <- sum(x$edge.length[ind]) / 2
  x$edge.length[ind] <- el
  x$node.label[1] = ""
  x
} 

for (i in 1:length(rooted_tree)){
  rooted_tree[[i]] <- pretty_root(rooted_tree[[i]])
}
```
\

We can plot the first tree with the bootstrap values (rounded to two digits) as node labels.

```{r plot_pars_trees, fig.width=16, fig.height=20}
par(mar = c(1,1,4,1))
plot(rooted_tree[[1]], type = "phy", edge.width=1) 
title(main="Parsimony tree")
nodelabels(round(as.numeric(rooted_tree[[1]]$node.label), digits = 2), frame = "none", adj=c(1,1), cex=.75)
add.scale.bar()
```
\
\

We could also plot all trees, to get a quick overview (in this case without the node labels):

```{r plot_pars_tree_wn, eval=F}
par(mar = c(1,1,4,1))
plot(rooted_tree, type = "phy", cex=.5, edge.width=2)
```
\

For this paper, we wrote the trees in *Newick*-format to edit them in *FigTree* [@figtree]. All trees can be written into one file.

```{r write_pars_trees, eval=F}
write.tree(tree_pars, file = "ParsimonyTrees.trees")
```
\

## Maximum Likelihood Trees
  
  
For maximum likelihood trees, the function `pml_bb` from *phangorn* can be used. For **unordered morphometric data**, `model="ER+ASC"` should be used. 

- **ER**: equal rates, all transitions occur at the same rate  
- **ASC**: ascertainment bias correction [@lewis2001likelihood] corrects for the fact that there are no constant sites (as is mostly the case with morphometric data); otherwise the branch lengths may be overestimated  

We calculate one tree with a strict clock (`method="ultrametric"`) and one with `method="unrooted"`. Ultrafast bootstrapping [@minh2013] values are calculated with this function automatically.
  
  
```{r ml_trees, cache=T}
# strict clock
ml_ULTRA <- pml_bb(morpho_pd, model="ER+ASC", rearrangement = "stochastic",
                       method="ultrametric",
                       control=pml.control(trace = 0))

# unrooted
ml <- pml_bb(morpho_pd, model="ER+ASC", rearrangement = "stochastic",
                 control=pml.control(trace = 0))
```
\

If we only want the tree object, we can get it with:
``` {r ml_bootstrap}
# strict clock
tree_ml_ULTRA <- ml_ULTRA$tree

# unrooted
tree_ml <- ml$tree
```
\

Rooting (or in the case of the ultrametric tree: rerooting) the tree can again be done with `root` and our function `pretty_root`, with the outgroup *Collohmannia johnstoni*.

```{r root_ml}
# strict clock
rooted_ml_ULTRA_tree <- root(tree_ml_ULTRA, "Collohmannia_johnstoni", resolve.root = TRUE,
                             edgelabel=TRUE)
rooted_ml_ULTRA_tree <- pretty_root(rooted_ml_ULTRA_tree)

# relaxed clock
rooted_ml_tree <- root(tree_ml, "Collohmannia_johnstoni", resolve.root = TRUE,
                             edgelabel=TRUE)
rooted_ml_tree <- pretty_root(rooted_ml_tree)
```
\

Then we can plot the two trees.

```{r plot_ml_trees, fig.width=16, fig.height=40}
par(mar = c(1,1,4,1), mfrow=2:1)
plot(rooted_ml_ULTRA_tree, type = "phy", edge.width=1) 
title(main = "Strict clock (ultrametric)")
nodelabels(round(as.numeric(rooted_ml_ULTRA_tree$node.label), digits = 2), frame = "none", adj=c(1,1), cex=.75)
add.scale.bar()

plot(rooted_ml_ULTRA_tree, type = "phy", edge.width=1) 
title(main = "Unrooted method")
nodelabels(round(as.numeric(rooted_ml_ULTRA_tree$node.label), digits = 2), frame = "none", adj=c(1,1), cex=.75)
add.scale.bar()
```
\
\

For this paper, we also wrote those trees in *Newick*-format, to edit them in *FigTree* [@figtree].
  
  
```{r write_ml_newick, eval=F}
write.tree(tree_ml_ULTRA, file = "MLTree_ULTRA.tree")
write.tree(tree_ml, file = "MLTree.tree")
```
\

## Bayesian inference

The bayesian trees were inferred with BEAST2 [@drummond2012bayesian; @bouckaert2019beast] using the *morph-models* package [@lewis2001likelihood]. The *nexus*-file was loaded into *BEAUti* and most of the default settings were used. Chain length was set to 50 million, trace and tree log every 1000. The random seed was set to 102.
Again, we used two clock methods: 

- strict clock [@zuckerkandl1965molecules]
- relaxed clock [@drummond2006relaxed]

The trace was later inspected in *Tracer* [@rambaut2018posterior] and the *.trees* files were summarized using the *TreeAnnotator* tool (maximum clade credibility tree with 0.5 posterior probability limit, median height nodes and 25% burn-in). For visualising *FigTree* [@figtree] was used.
\
\

## Session Info

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

## References